---
title: "Findings of 1990"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library("MASS")
library("plyr")
library("dplyr")
library("ggplot2")
library("knitr")
library("kableExtra")
```


The following variables are selected to see if the difference in income between men and women in 1990 vary depending on them: <br>

1. race<br>
2. age.start.alcohol.1984<br>
3. womens.roles.q1<br>
4. womens.roles.q2<br>
5. family.size.1979<br>


```{r}

nlsy <- read.csv("nlsy79_income.csv", header=TRUE)

# Change column names (mostly) to question name abbreviations
colnames(nlsy) <- c("VERSION_R25_2012",
    "CASEID_1979",
    "FAM-2A_1979",
    "FAM-POB_1979",
    "FAM-3_1979",
    "FAM-3A_1979",
    "FAM-RES_1979",
    "FAM-6_1979",
    "R_REL-1_COL_1979",
    "SCHOOL-31_1979",
    "MIL-6_1979",
    "womens.roles.q1",
    "womens.roles.q2",
    "womens.roles.q3",
    "womens.roles.q4",
    "womens.roles.q6",
    "womens.roles.q7",
    "womens.roles.q8",
    "EXP-OCC_1979",
    "EXP-9_1979",
    "race",
    "gender",
    "marital.status.1979",
    "family.size.1979",
    "POVSTATUS_1979",
    "POLICE-1_1980",
    "POLICE-1C_1980",
    "POLICE-2_1980",
    "age.start.alcohol.1984",
    "times.drug.use.1984",
    "age.start.drug.use.1984",
    "income.1990",
    "POVSTATUS_1990",
    "HGCREV90_1990",
    "JOBSNUM_1990",
    "household.children.number.1990",
    "AGEYCH90_1990",
    "DS-12_1998",
    "age.start.drug.use.1998",
    "industry.2000",
    "CPSOCC80.01_2000",
    "OCCSP-55I_CODE_2000",
    "Q2-15B_2000",
    "Q10-2_2000",
    "income.2000",
    "family.size.2000",
    "TNFI_TRUNC_2000",
    "POVSTATUS_2000",
    "MARSTAT-COL_2000",
    "marital.status.2000",
    "MO1M1B_XRND",
    "Q2-10B~Y_2012",
    "industry.2012",
    "OCCALL-EMP.01_2012",
    "OCCSP-55I_CODE_2012",
    "Q2-15A_2012",
    "Q12-6_2012",
    "income.2012",
    "estimated.low.income.2012",
    "estimated.high.income.2012",
    "Q13-18_TRUNC_2012",
    "Q13-18_SR000001_TRUNC_2012",
    "family.size.2012",
    "residence.region.2012",
    "highest.education.2012",
    "residence.urban.rural.2012",
    "JOBSNUM_2012")

# Map all negative values to missing (NA)
nlsy[nlsy < 0] <- NA

# Transform and relabel gender and race variables
nlsy = transform(nlsy, gender = as.factor(mapvalues(gender, 
                                                        c(1,2), 
                                                        c("Male", "Female"))))
nlsy = transform(nlsy, race = as.factor(mapvalues(race, 
                                                      c(1,2,3), 
                                                      c("Hispanic", "Black", "Other"))))

# See how many values are missing in each variables
empty.list = list()
for(i in 1:67){
  empty.list = append(empty.list, length(which(is.na(nlsy[i]))))
}
names(empty.list) = names(nlsy)

```


```{r}
# Transform relabel gender and race variables
nlsy = transform(nlsy, gender = as.factor(mapvalues(gender, 
                                                        c(1,2), 
                                                        c("Male", 
                                                          "Female"))))
nlsy = transform(nlsy, race = as.factor(mapvalues(race, 
                                                      c(1,2,3), 
                                                      c("Hispanic", 
                                                        "Black", 
                                                        "Other"))))
```

```{r}
# Transform the factors inside the residence.region.2012 variable to make it easier to understand
nlsy$residence.region.2012  = as.factor(mapvalues(nlsy$residence.region.2012 ,
                                                           c(1,2,3,4),
                                                           c("Northeast",
                                                             "North Central",
                                                             "South",
                                                             "West")))
```

```{r}
# Transform the factors inside the residence.urban.rural.2012 variable to make it easier to understand
nlsy$residence.urban.rural.2012  = as.factor(mapvalues(nlsy$residence.urban.rural.2012 ,
                                                           c(0,1,2),
                                                           c("Rural",
                                                             "Urban",
                                                             "Unknown"
                                                             )))
```

```{r}
# Transform the factors inside the marital.status.2000 variable to make it easier to understand
nlsy$marital.status.2000  = as.factor(mapvalues(nlsy$marital.status.2000 ,
                                                           c(0,1,2,3,6),
                                                           c("Never Married",
                                                             "Married",
                                                             "Separated",
                                                             "Divorced",
                                                             "Widowed")))
```

```{r}
# Modify and transform the industry.2000 variable to make it easier to analyse
for(i in 1:length(nlsy$industry.2000)) {
  if(is.na(nlsy[i, "industry.2000"])) {}
  else if(nlsy[i, "industry.2000"] >= 17
     && nlsy[i, "industry.2000"] <= 28) {
          nlsy[i, "industry.2000"] = 1
  }  else if(nlsy[i, "industry.2000"] >= 47
     && nlsy[i, "industry.2000"] <= 57) {
          nlsy[i, "industry.2000"] = 2
     } else if(nlsy[i, "industry.2000"] >= 67
     && nlsy[i, "industry.2000"] <= 77) {
          nlsy[i, "industry.2000"] = 3
     } else if(nlsy[i, "industry.2000"] >= 107
     && nlsy[i, "industry.2000"] <= 398) {
          nlsy[i, "industry.2000"] = 4
     } else if(nlsy[i, "industry.2000"] >= 407
     && nlsy[i, "industry.2000"] <= 479) {
          nlsy[i, "industry.2000"] = 5
     } else if(nlsy[i, "industry.2000"] >= 507
     && nlsy[i, "industry.2000"] <= 698) {
          nlsy[i, "industry.2000"] = 6
     } else if(nlsy[i, "industry.2000"] >= 707
     && nlsy[i, "industry.2000"] <= 718) {
          nlsy[i, "industry.2000"] = 7
     } else if(nlsy[i, "industry.2000"] >= 727
     && nlsy[i, "industry.2000"] <= 759) {
          nlsy[i, "industry.2000"] = 8
     } else if(nlsy[i, "industry.2000"] >= 769
     && nlsy[i, "industry.2000"] <= 798) {
          nlsy[i, "industry.2000"] = 9
     } else if(nlsy[i, "industry.2000"] >= 807
     && nlsy[i, "industry.2000"] <= 809) {
          nlsy[i, "industry.2000"] = 10
     } else if(nlsy[i, "industry.2000"] >= 828
     && nlsy[i, "industry.2000"] <= 897) {
          nlsy[i, "industry.2000"] = 11
     } else if(nlsy[i, "industry.2000"] >= 907
     && nlsy[i, "industry.2000"] <= 937) {
          nlsy[i, "industry.2000"] = 12
     }else{
       nlsy[i, "industry.2000"] = 13
     }
}

nlsy$industry.2000 = as.factor(
  mapvalues(
    nlsy$industry.2000,
    c(1,2,3,4,5,6,7,8,9,10,11,12,13),
    c("AGRICULTURE,FORESTRY AND FISHERIES",
      "MINING",
      "CONSTRUCTION",
      "MANUFACTURING",
      "TRANSPORTATION,COMMUNICATION,PUBLIC UTILITIES",
      "WHOLESALE AND RETAIL TRADE",
      "FINANCE,INSURANCE AND REAL ESTATE",
      "BUSINESS AND REPAIR SERVICES",
      "PERSONAL SERVICES",
      "ENTERTAINMENT AND RECREATION SERVICES",
      "PROFESSIONAL AND RELATED SERVICES",
      "PUBLIC ADMINISTRATION",
      "OTHERS")))

```

```{r}
# modify the industry.2012 variable to make it easier to analyse
for(i in 1:length(nlsy$industry.2012)) {
  if(is.na(nlsy[i, "industry.2012"])) {}
  else if(nlsy[i, "industry.2012"] >= 170
     && nlsy[i, "industry.2012"] <= 290) {
          nlsy[i, "industry.2012"] = 1
  }  else if(nlsy[i, "industry.2012"] >= 370
     && nlsy[i, "industry.2012"] <= 490) {
          nlsy[i, "industry.2012"] = 2
     } else if(nlsy[i, "industry.2012"] >= 570
     && nlsy[i, "industry.2012"] <= 690) {
          nlsy[i, "industry.2012"] = 3
     } else if(nlsy[i, "industry.2012"] == 770) {
          nlsy[i, "industry.2012"] = 4
     } else if(nlsy[i, "industry.2012"] >= 1070
     && nlsy[i, "industry.2012"] <= 3990) {
          nlsy[i, "industry.2012"] = 5
     } else if(nlsy[i, "industry.2012"] >= 4070
     && nlsy[i, "industry.2012"] <= 4590) {
          nlsy[i, "industry.2012"] = 6
     } else if(nlsy[i, "industry.2012"] >= 4670
     && nlsy[i, "industry.2012"] <= 5790) {
          nlsy[i, "industry.2012"] = 7
     } else if(nlsy[i, "industry.2012"] >= 6070
     && nlsy[i, "industry.2012"] <= 6390) {
          nlsy[i, "industry.2012"] = 8
     } else if(nlsy[i, "industry.2012"] >= 6470
     && nlsy[i, "industry.2012"] <= 6780) {
          nlsy[i, "industry.2012"] = 9
     } else if(nlsy[i, "industry.2012"] >= 6870
     && nlsy[i, "industry.2012"] <= 6990) {
          nlsy[i, "industry.2012"] = 10
     } else if(nlsy[i, "industry.2012"] >= 7070
     && nlsy[i, "industry.2012"] <= 7190) {
          nlsy[i, "industry.2012"] = 11
     } else if(nlsy[i, "industry.2012"] >= 7270
     && nlsy[i, "industry.2012"] <= 7490) {
          nlsy[i, "industry.2012"] = 12
     } else if(nlsy[i, "industry.2012"] >= 7570
     && nlsy[i, "industry.2012"] <= 7790) {
          nlsy[i, "industry.2012"] = 13
     } else if(nlsy[i, "industry.2012"] >= 7860
     && nlsy[i, "industry.2012"] <= 7890) {
          nlsy[i, "industry.2012"] = 14
     } else if(nlsy[i, "industry.2012"] >= 7970
     && nlsy[i, "industry.2012"] <= 8470) {
          nlsy[i, "industry.2012"] = 15
     } else if(nlsy[i, "industry.2012"] >= 8560
     && nlsy[i, "industry.2012"] <= 8590) {
          nlsy[i, "industry.2012"] = 16
     } else if(nlsy[i, "industry.2012"] >= 8660
     && nlsy[i, "industry.2012"] <= 8690) {
          nlsy[i, "industry.2012"] = 17
     } else if(nlsy[i, "industry.2012"] >= 8770
     && nlsy[i, "industry.2012"] <= 9290) {
          nlsy[i, "industry.2012"] = 18
     } else if(nlsy[i, "industry.2012"] >= 9370
     && nlsy[i, "industry.2012"] <= 9870) {
          nlsy[i, "industry.2012"] = 19
     } else if(nlsy[i, "industry.2012"] == 9890) {
          nlsy[i, "industry.2012"] = 20
     } else if(nlsy[i, "industry.2012"] >= 9950) {
          nlsy[i, "industry.2012"] = 21
     } else if(nlsy[i, "industry.2012"] >= 9990) {
          nlsy[i, "industry.2012"] = 22
     }  else {
          nlsy[i, "industry.2012"] = 25
     }
}

nlsy$industry.2012 = as.factor(
  mapvalues(
    nlsy$industry.2012,
    c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,25),
    c("Agriculture, Forestry, Fishing, and Hunting",
      "Mining",
      "Utilities",
      "Construction",
      "Manufacturing",
      "Wholesale Trade",
      "Retail Trade",
      "Transportation and Warehousing",
      "Information",
      "Finance and Insurance",
      "Real Estate",
      "Professional and Technical",
      "Administration and Management",
      "Education",
      "Health Care",
      "Arts and Entertainment",
      "Accomodation and Food",
      "Other services",
      "Public Administration",
      "Armed Forces",
      "Not in Labor Force",
      "Uncodeable",
      "Uncategorized")))
```

```{r}
# Transform womens.roles.q7 variable
nlsy$womens.roles.q7 = as.factor(mapvalues(nlsy$womens.roles.q7, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q1 variable
nlsy$womens.roles.q1 = as.factor(mapvalues(nlsy$womens.roles.q1, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q2 variable
nlsy$womens.roles.q2 = as.factor(mapvalues(nlsy$womens.roles.q2, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q4 variable
nlsy$womens.roles.q4 = as.factor(mapvalues(nlsy$womens.roles.q4, 
                                           c(1,2,3,4), 
                                           c("Strongly disagree", 
                                             "Disagree", 
                                             "Agree", 
                                             "Strongly agree")))
```

```{r}

for(i in 1:12686){
  if(is.na(nlsy[i, "income.2012"]) & 
     !is.na(nlsy[i, "estimated.low.income.2012"]) &
     !is.na(nlsy[i, "estimated.high.income.2012"])) {
    nlsy[i, "income.2012"] = 
      (nlsy[i, "estimated.low.income.2012"] + nlsy[i, "estimated.high.income.2012"]) / 2
  }
}

# Remove all entries that miss any income
cleaned.nlsy = nlsy[which(!is.na(nlsy$income.1990) 
                          & !is.na(nlsy$income.2000) 
                          & !is.na(nlsy$income.2012)), ]

```


<h1>T-Test</h1>
Use t.test to test if there is significant difference in income between male and female in 1990. 

```{r}

income1990.t.test = t.test(income.1990 ~ gender, cleaned.nlsy)
income1990.t.test

```

**The outcome shows that the average different in income between male and female in 1990 is `r round(income1990.t.test$estimate[2] - income1990.t.test$estimate[1], 0)` dollar, with the p value `r round(income1990.t.test$p.value, 10)`, which is very significant. **



<h1>Model 1</h1>
In the first model, only the data of those whose income is more than 0 but not top coded is used to do linear regression. The whole cleaned data frame is divided into two parts, one male and one female, and regression seperately. 

```{r}
clean.nlsy.income1990.1 = select(cleaned.nlsy, 
                                 income.1990, 
                                 gender,
                                 race, 
                                 age.start.alcohol.1984, 
                                 womens.roles.q1,
                                 womens.roles.q2,
                                 family.size.1979
                                 ) %>%
  filter(income.1990>0, 
         income.1990<74283) 

```


Test if income is normally distributed using qqplot. 
```{r}
qqnorm(clean.nlsy.income1990.1$income.1990)
qqline(clean.nlsy.income1990.1$income.1990)
```

**It looks like the data is not normally distributed. **



Divide the data frame into 2 parts according to gender.
```{r}
clean.nlsy.income1990.1.male = filter(clean.nlsy.income1990.1, gender == "Male") %>%
  select(-gender)
clean.nlsy.income1990.1.female = filter(clean.nlsy.income1990.1, gender == "Female")%>%
  select(-gender)
```


First test if there are collinear variables in male.
```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y, use = "complete.obs"))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = pmax(1, cex.cor * r))
}
pairs(clean.nlsy.income1990.1.male[, c("race", 
                                       "age.start.alcohol.1984", 
                                       "womens.roles.q1", 
                                       "womens.roles.q2", 
                                       "family.size.1979")], 
      lower.panel = panel.cor)
```

**From the plot, it can be seen that no variables are collinear in male. **



Then test female. 
```{r}
pairs(clean.nlsy.income1990.1.female[, c("race", 
                                         "age.start.alcohol.1984", 
                                         "womens.roles.q1", 
                                         "womens.roles.q2", 
                                         "family.size.1979")], 
      lower.panel = panel.cor)
```

**From the plot, no variables are collinear in female. **



So it it safe to do linear regression. 
<br>Regression summary of male: 

```{r}
clean.nlsy.income1990.1.male.lm = lm(income.1990 ~ ., clean.nlsy.income1990.1.male)
summary(clean.nlsy.income1990.1.male.lm)$coef %>%
  kable() %>%
  kable_styling()
```
<br> Check if the model is good. 
```{r}
par(mfrow=c(2,2))
plot(clean.nlsy.income1990.1.male.lm)
```

**From the 4 plots, it can be seen that the model is quite good because: ** <br>
**In the 1st plot, data is distributed randomly; **<br>
**In the 2nd plot, data is nearly a straight line from the lower left corner to the upper right corner; **<br>
**In the 3rd plot, the red line is flat; **<br>
**In the 4th plot, very few data exceeds 3 or -3; **<br>



Regression summary of female: 
```{r}
clean.nlsy.income1990.1.female.lm = lm(income.1990 ~ ., clean.nlsy.income1990.1.female)
summary(clean.nlsy.income1990.1.female.lm)$coef %>%
  kable() %>%
  kable_styling()
```
<br>Check if the model is good. 
```{r}
par(mfrow=c(2,2))
plot(clean.nlsy.income1990.1.female.lm)
```

**From the 4 plots, the model is quite good for similar reasons**
<br>

<br>
**Significant variables (with p value <= 0.05) of male are shown below: **
```{r}
summary(clean.nlsy.income1990.1.male.lm)$coef[which(
  summary(clean.nlsy.income1990.1.male.lm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```

<br>
**Significant variables (with p value <= 0.05) of female are shown below: **
```{r}
summary(clean.nlsy.income1990.1.female.lm)$coef[which(
  summary(clean.nlsy.income1990.1.female.lm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```
<br>
**From the 2 tables, it is found that the difference in income in 1990 vary depending on: **<br>
**1. Race. The income of people of different races have significant different income. The income of Hispanic male is `r summary(clean.nlsy.income1990.1.male.lm)$coef[2, 1]` more than Black male which is the base line, while that of Hispanic female is only `r summary(clean.nlsy.income1990.1.female.lm)$coef[2, 1]` more than Black female. It means that difference in income of Hispanic is `r summary(clean.nlsy.income1990.1.male.lm)$coef[2, 1] - summary(clean.nlsy.income1990.1.female.lm)$coef[2, 1]` more than difference of Black. Similarly, difference of Other races is `r summary(clean.nlsy.income1990.1.male.lm)$coef[3, 1] - summary(clean.nlsy.income1990.1.female.lm)$coef[3, 1]` more. **<br>
**2. First age of alcohol consumption. This variable can significantly influence female with the p value `r summary(clean.nlsy.income1990.1.female.lm)$coef[4,4]`, and every year later of starting drinking, `r summary(clean.nlsy.income1990.1.female.lm)$coef[4,1]` more income she will have. But this variable of male is not significant, only with the p value `r summary(clean.nlsy.income1990.1.male.lm)$coef[4,4]` .**<br>
**3. Whether the responder agrees that a wife who carries out her full family responsibilities does not have time for outside employment (women.role.2). The difference between those who disagree narrows `r summary(clean.nlsy.income1990.1.female.lm)$coef[8, 1] - summary(clean.nlsy.income1990.1.male.lm)$coef[8, 1]`. The difference between those who strongly disagree is even more, which is `r summary(clean.nlsy.income1990.1.female.lm)$coef[10, 1] - summary(clean.nlsy.income1990.1.male.lm)$coef[10, 1]`. ** <br>
**4. Family size. This variable can significantly influence male with the p value `r summary(clean.nlsy.income1990.1.male.lm)$coef[11,4]`, and every one more person in family, `r -summary(clean.nlsy.income1990.1.male.lm)$coef[11,1]` less income he will have. But this variable of female is not significant, only with the p value `r summary(clean.nlsy.income1990.1.female.lm)$coef[11,4]`. **<br>
<br>



<h1>Model 2</h1>
In the second model, data of those whose income is more than 0 is used to create a logistic variable of whether the responder is of high income (whether the responder’s income is top coded), and do logistic regression.

```{r}
clean.nlsy.income1990.2 = select(cleaned.nlsy, 
                                 income.1990, 
                                 gender,
                                 race, 
                                 age.start.alcohol.1984, 
                                 womens.roles.q1,
                                 womens.roles.q2,
                                 family.size.1979
                                 ) %>%
  filter(income.1990>0) 

for(i in 1:4999) {
  if(clean.nlsy.income1990.2[i, "income.1990"] == 74283) {
    clean.nlsy.income1990.2[i, "high.income"] = TRUE
  } else{
    clean.nlsy.income1990.2[i, "high.income"] = FALSE
  }
}

clean.nlsy.income1990.2 = select(clean.nlsy.income1990.2, -income.1990)

clean.nlsy.income1990.2.male = filter(clean.nlsy.income1990.2, gender == "Male") %>%
  select(-gender)
clean.nlsy.income1990.2.female = filter(clean.nlsy.income1990.2, gender == "Female")%>%
  select(-gender)
```


Regression summary of male: 
```{r}
clean.nlsy.income1990.2.male.glm = lm(high.income ~ ., 
                                      family = binomial(), 
                                      data = clean.nlsy.income1990.2.male)
summary(clean.nlsy.income1990.2.male.glm)$coef %>%
  kable() %>%
  kable_styling()
```



Regression summary of female: 
```{r}
clean.nlsy.income1990.2.female.glm = glm(high.income ~ ., 
                                         family = binomial(),
                                         data = clean.nlsy.income1990.2.female)
summary(clean.nlsy.income1990.2.female.glm)$coef %>%
  kable() %>%
  kable_styling()
```

**Significant variables (with p value <= 0.05) of male are shown below:**

```{r}
summary(clean.nlsy.income1990.2.male.glm)$coef[which(
  summary(clean.nlsy.income1990.2.male.glm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```

**There are no significant variables (with p value <= 0.05) of female**<br><br>

**It is found that whether one has top 2% income in 1990 vary depending on: **<br>
**1. Race. For male of other race compared to black, the odds of having an income of top 2% increase by a factor of `r exp(summary(clean.nlsy.income1990.2.male.glm)$coef[3,1])` with a p value of `r (summary(clean.nlsy.income1990.2.male.glm)$coef[3,4])`, while there is no significant relationship between female of other race and black female. **<br>
**2. Whether the responder agrees that a wife who carries out her full family responsibilities does not have time for outside employment (women.role.2). For male who strongly disagree with it compared to male who agree with it, the odds of having an income of top 2% increase by a factor of `r exp(summary(clean.nlsy.income1990.2.male.glm)$coef[10,1])` with a p value of `r (summary(clean.nlsy.income1990.2.male.glm)$coef[10,4])`, while there is no significant relationship for that of female. **<br>


<h1>Model 3</h1>
In the third model, every data is used to create a logistic variable of whether the responder has income (whether the income is more than 0), and do logistic regression.

```{r}
clean.nlsy.income1990.3 = select(cleaned.nlsy, 
                                 income.1990, 
                                 gender,
                                 race, 
                                 age.start.alcohol.1984, 
                                 womens.roles.q1,
                                 womens.roles.q2,
                                 family.size.1979
                                 ) 

for(i in 1:6021) {
  if(clean.nlsy.income1990.3[i, "income.1990"] == 0) {
    clean.nlsy.income1990.3[i, "have.income"] = FALSE
  } else{
    clean.nlsy.income1990.3[i, "have.income"] = TRUE
  }
}

clean.nlsy.income1990.3 = select(clean.nlsy.income1990.3, -income.1990)

clean.nlsy.income1990.3.male = filter(clean.nlsy.income1990.3, gender == "Male") %>%
  select(-gender)
clean.nlsy.income1990.3.female = filter(clean.nlsy.income1990.3, gender == "Female")%>%
  select(-gender)
```



Regression summary of male: 
```{r}
clean.nlsy.income1990.3.male.glm = lm(have.income ~ ., 
                                      family = binomial(), 
                                      data = clean.nlsy.income1990.3.male)
summary(clean.nlsy.income1990.3.male.glm)$coef %>%
  kable() %>%
  kable_styling()
```



Regression summary of female: 
```{r}
clean.nlsy.income1990.3.female.glm = glm(have.income ~ ., 
                                         family = binomial(),
                                         data = clean.nlsy.income1990.3.female)
summary(clean.nlsy.income1990.3.female.glm)$coef %>%
  kable() %>%
  kable_styling()
```

**Significant variables (with p value <= 0.05) of male are shown below: **

```{r}
summary(clean.nlsy.income1990.3.male.glm)$coef[which(
  summary(clean.nlsy.income1990.3.male.glm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```

**Significant variables (with p value <= 0.05) of female are shown below: **

```{r}
summary(clean.nlsy.income1990.3.female.glm)$coef[which(
  summary(clean.nlsy.income1990.3.female.glm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```

**It is found that whether one has income in 1990 vary depending on: **<br>
**1. Race. For male of hispanic race compared to black, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income1990.3.male.glm)$coef[2,1])` with a p value of `r round(summary(clean.nlsy.income1990.3.male.glm)$coef[2,4])`, but there is no such significant relationship of female. For male of other race compared to black, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income1990.3.male.glm)$coef[3,1])` with a p value of `r (summary(clean.nlsy.income1990.3.male.glm)$coef[3,4])`. For female of other race compared to black, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income1990.3.female.glm)$coef[3,1])` with a p value of `r (summary(clean.nlsy.income1990.3.female.glm)$coef[3,4])`. **<br>
**2. Whether the responder agrees that a woman's place is in the home, not in the office or shop. For female who disagree compared to agree, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income1990.3.female.glm)$coef[5,1])` with a p value of `r (summary(clean.nlsy.income1990.3.female.glm)$coef[5,4])`. For female who strongly disagree compared to agree, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income1990.3.female.glm)$coef[7,1])` with a p value of `r round(summary(clean.nlsy.income1990.3.female.glm)$coef[7,4], 5)`. However, there is no significant relationship of male. **<br>
**3. First age of alcohol consumption. For evary one year later a male starts to drink, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income1990.3.male.glm)$coef[4,1])` with a p value of `r (summary(clean.nlsy.income1990.3.male.glm)$coef[4,4])`. There is no such significant relationship in female. **<br><br>


<h1>Interaction Model</h1>


Here we use interaction model to study how the difference in income between male and female varies depending on other factors. Three different models are introduced to respectively depict three income variables in different years.


For the income in 1990, all the variables mentioned above related to year 1990 are taken into consideration here at first. We select the variabels needed into `clean.nlsy.income1990.4`. For the income of 1990, we exclude the top 2% and zero income data.
```{r}
clean.nlsy.income1990.4 = select(cleaned.nlsy, 
                                 income.1990,
                                 gender,
                                 womens.roles.q1,
                                 womens.roles.q2,
                                 race,
                                 family.size.1979
                                 ) %>%
  filter(income.1990 > 0, 
         income.1990 < max(cleaned.nlsy$income.1990))
```


To test whether the income.1990 in clean.nlsy.income1990.4 is normally distributed, `qqmorm()` and `qqline()` are used here.
```{r}
par(mfrow=c(1,2))
qqnorm(clean.nlsy.income1990.4$income.1990^0.5)
qqline(clean.nlsy.income1990.4$income.1990^0.5)
qqnorm(clean.nlsy.income1990.4$income.1990)
qqline(clean.nlsy.income1990.4$income.1990) 
```

From the figure above, `income.1990^0.5` (left one) is more normally distributed than `income.1990` (right one). Therefore, `income.1990^0.5` is used in this interaction regression model.

Make each variable interact with `gender`. The summary and plots are as follows:
```{r}
income.lm.interact.1990 = lm(income.1990^0.5 ~ 
                               womens.roles.q1*gender +
                               womens.roles.q2*gender +
                               race*gender + 
                               family.size.1979*gender, 
                             data = clean.nlsy.income1990.4)
summary(income.lm.interact.1990)
par(mfrow=c(2,2))
plot(income.lm.interact.1990)
```

We find that the interaction term between gender and `womens.roles.q1` is not significant, but the p-value shows that the `womens.roles.q1` value itself is statistically significant. Therefore, we remove the interaction term between gender and `womens.roles.q1`. The results are as follows:

```{r}
summary(income.lm.interact.1990)
par(mfrow=c(2,2))
plot(income.lm.interact.1990)
```

Be careful that the dependant variable here is `income.1990^0.5` rather than `income.1990`, so the interperation of the coefficients will be different.

**According to the results, generally, the `income.1990^0.5` of male will be `r round(income.lm.interact.1990$coefficients["genderMale"], 2)` higher than female. Therefore, the income difference between male and female in 1990 is `r round((summary(income.lm.interact.1990)[[4]][1] + income.lm.interact.1990$coefficients["genderMale"])^2 - (summary(income.lm.interact.1990)[[4]][1])^2, 2)`**<br>

There are two interaction terms indicate the significant association to the difference of income between male and female.


+ `genderMale:raceOther`: the **Other race** significantly incluence the income difference between male and femal. In "Other" race, the **difference** of `income.1990^0.5` between male and femal increases `r round(income.lm.interact.1990$coefficients["genderMale:raceOther"], 2)`. Therefore, the income difference between male and femal in "Other" race will increase `r round((summary(income.lm.interact.1990)[[4]][1] + income.lm.interact.1990$coefficients["genderMale"] + income.lm.interact.1990$coefficients["raceOther"] + income.lm.interact.1990$coefficients["genderMale:raceOther"])^2 - (summary(income.lm.interact.1990)[[4]][1] + income.lm.interact.1990$coefficients["raceOther"])^2 - ((summary(income.lm.interact.1990)[[4]][1] + income.lm.interact.1990$coefficients["genderMale"])^2 - (summary(income.lm.interact.1990)[[4]][1])^2),2)`.

+ `genderMale:womens.roles.q2Strongly disagree`: for male and female who **Strongly disagree** with the question "Wife with family has no time for other employment?", their difference of `income.1990^0.5` decreases `r -round(income.lm.interact.1990$coefficients["genderMale:womens.roles.q2Strongly disagree"], 2)`.

+ `genderMale:womens.roles.q2Disagree`: for male and female who **Disagree** with the question "Wife with family has no time for other employment?", their difference of `income.1990^0.5` decreases `r -round(income.lm.interact.1990$coefficients["genderMale:womens.roles.q2Disagree"], 2)`.




