---
title: "Findings of 2012"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library("MASS")
library("plyr")
library("dplyr")
library("ggplot2")
library("knitr")
library("kableExtra")
```

The following variables are selected to see if the difference in income between men and women in 2012 vary depending on them: <br>

1. race<br> 
2. number.drink.per.day.2012<br>
3. residence.region.2012<br>
4. highest.education.2012<br>
5. family.size.2012<br>
6. industry.2012


```{r}
nlsy <- read.csv("nlsy79_income.csv", header=TRUE)

# Change column names (mostly) to question name abbreviations
colnames(nlsy) <- c("VERSION_R25_2012",
    "CASEID_1979",
    "FAM-2A_1979",
    "FAM-POB_1979",
    "FAM-3_1979",
    "FAM-3A_1979",
    "FAM-RES_1979",
    "FAM-6_1979",
    "R_REL-1_COL_1979",
    "SCHOOL-31_1979",
    "MIL-6_1979",
    "womens.roles.q1",
    "womens.roles.q2",
    "womens.roles.q3",
    "womens.roles.q4",
    "womens.roles.q6",
    "womens.roles.q7",
    "womens.roles.q8",
    "EXP-OCC_1979",
    "EXP-9_1979",
    "race",
    "gender",
    "marital.status.1979",
    "family.size.1979",
    "POVSTATUS_1979",
    "POLICE-1_1980",
    "POLICE-1C_1980",
    "POLICE-2_1980",
    "age.start.alcohol.1984",
    "times.drug.use.1984",
    "age.start.drug.use.1984",
    "income.1990",
    "POVSTATUS_1990",
    "HGCREV90_1990",
    "JOBSNUM_1990",
    "household.children.number.1990",
    "AGEYCH90_1990",
    "DS-12_1998",
    "age.start.drug.use.1998",
    "industry.2000",
    "CPSOCC80.01_2000",
    "OCCSP-55I_CODE_2000",
    "Q2-15B_2000",
    "Q10-2_2000",
    "income.2000",
    "family.size.2000",
    "TNFI_TRUNC_2000",
    "POVSTATUS_2000",
    "MARSTAT-COL_2000",
    "marital.status.2000",
    "MO1M1B_XRND",
    "Q2-10B~Y_2012",
    "industry.2012",
    "OCCALL-EMP.01_2012",
    "OCCSP-55I_CODE_2012",
    "Q2-15A_2012",
    "number.drink.per.day.2012",
    "income.2012",
    "estimated.low.income.2012",
    "estimated.high.income.2012",
    "Q13-18_TRUNC_2012",
    "Q13-18_SR000001_TRUNC_2012",
    "family.size.2012",
    "residence.region.2012",
    "highest.education.2012",
    "residence.urban.rural.2012",
    "JOBSNUM_2012")

# Map all negative values to missing (NA)
nlsy[nlsy < 0] <- NA
```


```{r}
# Transform relabel gender and race variables
nlsy = transform(nlsy, gender = as.factor(mapvalues(gender, 
                                                        c(1,2), 
                                                        c("Male", 
                                                          "Female"))))
nlsy = transform(nlsy, race = as.factor(mapvalues(race, 
                                                      c(1,2,3), 
                                                      c("Hispanic", 
                                                        "Black", 
                                                        "Other"))))
```

```{r}
# Transform the factors inside the residence.region.2012 variable to make it easier to understand
nlsy$residence.region.2012  = as.factor(mapvalues(nlsy$residence.region.2012 ,
                                                           c(1,2,3,4),
                                                           c("Northeast",
                                                             "North Central",
                                                             "South",
                                                             "West")))
```

```{r}
# Transform the factors inside the residence.urban.rural.2012 variable to make it easier to understand
nlsy$residence.urban.rural.2012  = as.factor(mapvalues(nlsy$residence.urban.rural.2012 ,
                                                           c(0,1,2),
                                                           c("Rural",
                                                             "Urban",
                                                             "Unknown"
                                                             )))
```

```{r}
# Transform the factors inside the marital.status.2000 variable to make it easier to understand
nlsy$marital.status.2000  = as.factor(mapvalues(nlsy$marital.status.2000 ,
                                                           c(0,1,2,3,6),
                                                           c("Never Married",
                                                             "Married",
                                                             "Separated",
                                                             "Divorced",
                                                             "Widowed")))
```

```{r}
# Modify and transform the industry.2000 variable to make it easier to analyse
for(i in 1:length(nlsy$industry.2000)) {
  if(is.na(nlsy[i, "industry.2000"])) {}
  else if(nlsy[i, "industry.2000"] >= 17
     && nlsy[i, "industry.2000"] <= 28) {
          nlsy[i, "industry.2000"] = 1
  }  else if(nlsy[i, "industry.2000"] >= 47
     && nlsy[i, "industry.2000"] <= 57) {
          nlsy[i, "industry.2000"] = 2
     } else if(nlsy[i, "industry.2000"] >= 67
     && nlsy[i, "industry.2000"] <= 77) {
          nlsy[i, "industry.2000"] = 3
     } else if(nlsy[i, "industry.2000"] >= 107
     && nlsy[i, "industry.2000"] <= 398) {
          nlsy[i, "industry.2000"] = 4
     } else if(nlsy[i, "industry.2000"] >= 407
     && nlsy[i, "industry.2000"] <= 479) {
          nlsy[i, "industry.2000"] = 5
     } else if(nlsy[i, "industry.2000"] >= 507
     && nlsy[i, "industry.2000"] <= 698) {
          nlsy[i, "industry.2000"] = 6
     } else if(nlsy[i, "industry.2000"] >= 707
     && nlsy[i, "industry.2000"] <= 718) {
          nlsy[i, "industry.2000"] = 7
     } else if(nlsy[i, "industry.2000"] >= 727
     && nlsy[i, "industry.2000"] <= 759) {
          nlsy[i, "industry.2000"] = 8
     } else if(nlsy[i, "industry.2000"] >= 769
     && nlsy[i, "industry.2000"] <= 798) {
          nlsy[i, "industry.2000"] = 9
     } else if(nlsy[i, "industry.2000"] >= 807
     && nlsy[i, "industry.2000"] <= 809) {
          nlsy[i, "industry.2000"] = 10
     } else if(nlsy[i, "industry.2000"] >= 828
     && nlsy[i, "industry.2000"] <= 897) {
          nlsy[i, "industry.2000"] = 11
     } else if(nlsy[i, "industry.2000"] >= 907
     && nlsy[i, "industry.2000"] <= 937) {
          nlsy[i, "industry.2000"] = 12
     }else{
       nlsy[i, "industry.2000"] = 13
     }
}

nlsy$industry.2000 = as.factor(
  mapvalues(
    nlsy$industry.2000,
    c(1,2,3,4,5,6,7,8,9,10,11,12,13),
    c("AGRICULTURE,FORESTRY AND FISHERIES",
      "MINING",
      "CONSTRUCTION",
      "MANUFACTURING",
      "TRANSPORTATION,COMMUNICATION,PUBLIC UTILITIES",
      "WHOLESALE AND RETAIL TRADE",
      "FINANCE,INSURANCE AND REAL ESTATE",
      "BUSINESS AND REPAIR SERVICES",
      "PERSONAL SERVICES",
      "ENTERTAINMENT AND RECREATION SERVICES",
      "PROFESSIONAL AND RELATED SERVICES",
      "PUBLIC ADMINISTRATION",
      "OTHERS")))

```

```{r}
# modify the industry.2012 variable to make it easier to analyse
for(i in 1:length(nlsy$industry.2012)) {
  if(is.na(nlsy[i, "industry.2012"])) {}
  else if(nlsy[i, "industry.2012"] >= 170
     && nlsy[i, "industry.2012"] <= 290) {
          nlsy[i, "industry.2012"] = 1
  }  else if(nlsy[i, "industry.2012"] >= 370
     && nlsy[i, "industry.2012"] <= 490) {
          nlsy[i, "industry.2012"] = 2
     } else if(nlsy[i, "industry.2012"] >= 570
     && nlsy[i, "industry.2012"] <= 690) {
          nlsy[i, "industry.2012"] = 3
     } else if(nlsy[i, "industry.2012"] == 770) {
          nlsy[i, "industry.2012"] = 4
     } else if(nlsy[i, "industry.2012"] >= 1070
     && nlsy[i, "industry.2012"] <= 3990) {
          nlsy[i, "industry.2012"] = 5
     } else if(nlsy[i, "industry.2012"] >= 4070
     && nlsy[i, "industry.2012"] <= 4590) {
          nlsy[i, "industry.2012"] = 6
     } else if(nlsy[i, "industry.2012"] >= 4670
     && nlsy[i, "industry.2012"] <= 5790) {
          nlsy[i, "industry.2012"] = 7
     } else if(nlsy[i, "industry.2012"] >= 6070
     && nlsy[i, "industry.2012"] <= 6390) {
          nlsy[i, "industry.2012"] = 8
     } else if(nlsy[i, "industry.2012"] >= 6470
     && nlsy[i, "industry.2012"] <= 6780) {
          nlsy[i, "industry.2012"] = 9
     } else if(nlsy[i, "industry.2012"] >= 6870
     && nlsy[i, "industry.2012"] <= 6990) {
          nlsy[i, "industry.2012"] = 10
     } else if(nlsy[i, "industry.2012"] >= 7070
     && nlsy[i, "industry.2012"] <= 7190) {
          nlsy[i, "industry.2012"] = 11
     } else if(nlsy[i, "industry.2012"] >= 7270
     && nlsy[i, "industry.2012"] <= 7490) {
          nlsy[i, "industry.2012"] = 12
     } else if(nlsy[i, "industry.2012"] >= 7570
     && nlsy[i, "industry.2012"] <= 7790) {
          nlsy[i, "industry.2012"] = 13
     } else if(nlsy[i, "industry.2012"] >= 7860
     && nlsy[i, "industry.2012"] <= 7890) {
          nlsy[i, "industry.2012"] = 14
     } else if(nlsy[i, "industry.2012"] >= 7970
     && nlsy[i, "industry.2012"] <= 8470) {
          nlsy[i, "industry.2012"] = 15
     } else if(nlsy[i, "industry.2012"] >= 8560
     && nlsy[i, "industry.2012"] <= 8590) {
          nlsy[i, "industry.2012"] = 16
     } else if(nlsy[i, "industry.2012"] >= 8660
     && nlsy[i, "industry.2012"] <= 8690) {
          nlsy[i, "industry.2012"] = 17
     } else if(nlsy[i, "industry.2012"] >= 8770
     && nlsy[i, "industry.2012"] <= 9290) {
          nlsy[i, "industry.2012"] = 18
     } else if(nlsy[i, "industry.2012"] >= 9370
     && nlsy[i, "industry.2012"] <= 9870) {
          nlsy[i, "industry.2012"] = 19
     } else if(nlsy[i, "industry.2012"] == 9890) {
          nlsy[i, "industry.2012"] = 20
     } else if(nlsy[i, "industry.2012"] >= 9950) {
          nlsy[i, "industry.2012"] = 21
     } else if(nlsy[i, "industry.2012"] >= 9990) {
          nlsy[i, "industry.2012"] = 22
     }  else {
          nlsy[i, "industry.2012"] = 25
     }
}

nlsy$industry.2012 = as.factor(
  mapvalues(
    nlsy$industry.2012,
    c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,25),
    c("Agriculture, Forestry, Fishing, and Hunting",
      "Mining",
      "Utilities",
      "Construction",
      "Manufacturing",
      "Wholesale Trade",
      "Retail Trade",
      "Transportation and Warehousing",
      "Information",
      "Finance and Insurance",
      "Real Estate",
      "Professional and Technical",
      "Administration and Management",
      "Education",
      "Health Care",
      "Arts and Entertainment",
      "Accomodation and Food",
      "Other services",
      "Public Administration",
      "Armed Forces",
      "Not in Labor Force",
      "Uncodeable",
      "Uncategorized")))
```

```{r}
# Transform womens.roles.q7 variable
nlsy$womens.roles.q7 = as.factor(mapvalues(nlsy$womens.roles.q7, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q1 variable
nlsy$womens.roles.q1 = as.factor(mapvalues(nlsy$womens.roles.q1, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q2 variable
nlsy$womens.roles.q2 = as.factor(mapvalues(nlsy$womens.roles.q2, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q4 variable
nlsy$womens.roles.q4 = as.factor(mapvalues(nlsy$womens.roles.q4, 
                                           c(1,2,3,4), 
                                           c("Strongly disagree", 
                                             "Disagree", 
                                             "Agree", 
                                             "Strongly agree")))
```

```{r}
for(i in 1:12686){
  if(is.na(nlsy[i, "income.2012"]) & 
     !is.na(nlsy[i, "estimated.low.income.2012"]) &
     !is.na(nlsy[i, "estimated.high.income.2012"])) {
    nlsy[i, "income.2012"] = 
      (nlsy[i, "estimated.low.income.2012"] + nlsy[i, "estimated.high.income.2012"]) / 2
  }
}

# Remove all entries that miss any income
cleaned.nlsy = nlsy[which(!is.na(nlsy$income.1990) 
                          & !is.na(nlsy$income.2000) 
                          & !is.na(nlsy$income.2012)), ]
```


<h1>T-Test</h1>
Use t.test to test if there is significant difference in income between male and female in 2012. 

```{r}
income2012.t.test = t.test(income.2012 ~ gender, cleaned.nlsy)
income2012.t.test
```

**The outcome shows that the average different in income between male and female in 2012 is `r round(income2012.t.test$estimate[2] - income2012.t.test$estimate[1], 1)` dollar, with the p value `r round(income2012.t.test$p.value, 10)`**



<h1>Model 1</h1>
In the first model, data of those whose income is more than 0 but not top coded is used to do linear regression. We divide the whole cleaned data frame into two parts, one male and one female, and do regression seperately.

```{r}
clean.nlsy.income2012.1 = select(cleaned.nlsy, 
                                 income.2012, 
                                 gender,
                                 race,  
                                 number.drink.per.day.2012, 
                                 residence.region.2012,
                                 highest.education.2012,
                                 family.size.2012,
                                 industry.2012
                                 ) %>%
  filter(income.2012>0, 
         income.2012<343830)
```


First test if income is normally distributed using qqplot.
```{r}
qqnorm(clean.nlsy.income2012.1$income.2012)
qqline(clean.nlsy.income2012.1$income.2012)
```

**The income is not normally distributed. So should be dealt with. Doing square root is a good way.**


```{r}
clean.nlsy.income2012.1$income.2012 = clean.nlsy.income2012.1$income.2012^0.5
qqnorm(clean.nlsy.income2012.1$income.2012)
qqline(clean.nlsy.income2012.1$income.2012)
```

**Much better now. **


Divide the data frame into 2 by gender. 
```{r}
clean.nlsy.income2012.1.male = filter(clean.nlsy.income2012.1, gender == "Male") %>%
  select(-gender)
clean.nlsy.income2012.1.female = filter(clean.nlsy.income2012.1, gender == "Female")%>%
  select(-gender)
```


Then test if there are collinear variables. First male.
```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y, use = "complete.obs"))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = pmax(1, cex.cor * r))
}
pairs(clean.nlsy.income2012.1.male[, c("race", 
                                       "number.drink.per.day.2012", 
                                       "residence.region.2012", 
                                       "highest.education.2012", 
                                       "family.size.2012", 
                                       "industry.2012")], 
      lower.panel = panel.cor)
```

**There is no collinear**

Then female.

```{r}
pairs(clean.nlsy.income2012.1.female[, c("race", 
                                       "number.drink.per.day.2012", 
                                       "residence.region.2012", 
                                       "highest.education.2012", 
                                       "family.size.2012", 
                                       "industry.2012")], 
      lower.panel = panel.cor)
```

**There is no collinear**


Now can do linear regression. <br>
Regression summary of male: 

```{r}
clean.nlsy.income2012.1.male.lm = lm(income.2012 ~ ., clean.nlsy.income2012.1.male)
summary(clean.nlsy.income2012.1.male.lm)$coef %>%
  kable() %>%
  kable_styling()
```

```{r}
par(mfrow=c(2,2))
plot(clean.nlsy.income2012.1.male.lm)
```

**From the 4 plots, we can see that the model is quite good except for a few data exceeds -3 in plot 4.**


Regression summary of female: 
```{r}
clean.nlsy.income2012.1.female.lm = lm(income.2012 ~ ., clean.nlsy.income2012.1.female)
summary(clean.nlsy.income2012.1.female.lm)$coef %>%
  kable() %>%
  kable_styling()
```

```{r}
par(mfrow=c(2,2))
plot(clean.nlsy.income2012.1.female.lm)
```

**From the 4 plots, we can see that the model is quite good except for some data in plot 4 exceeds 3 or -3.**
<br><br>
**Significant variables (with p value <= 0.05) of male are shown below: **

```{r}
summary(clean.nlsy.income2012.1.male.lm)$coef[which(
  summary(clean.nlsy.income2012.1.male.lm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```
<br>
**Significant variables (with p value <= 0.05) of female are shown below: **

```{r}
summary(clean.nlsy.income2012.1.female.lm)$coef[which(
  summary(clean.nlsy.income2012.1.female.lm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```

<br>**From the 2 tables, it is found that the difference in income in 2000 vary depending on: **<br>
**1. Race. The income after square root of hispanic male is `r summary(clean.nlsy.income2012.1.male.lm)$coef[2,1]` more than that of black male with a p value of `r summary(clean.nlsy.income2012.1.male.lm)$coef[2,4]`, while The income after square root of hispanic female is `r summary(clean.nlsy.income2012.1.female.lm)$coef[2,1]` more than that of black female with a p value of `r summary(clean.nlsy.income2012.1.female.lm)$coef[2,4]`, so there is `r summary(clean.nlsy.income2012.1.male.lm)$coef[2,1] - summary(clean.nlsy.income2012.1.female.lm)$coef[2,1]` more difference. The income after square root of male of other race is `r summary(clean.nlsy.income2012.1.male.lm)$coef[3,1]` more than that of black female with a p value of `r summary(clean.nlsy.income2012.1.male.lm)$coef[3,4]`. However, there is no significant in difference between other female and black female in income after square root. **<br>
**2. Number of drink per day. For male, every one more drink per day, the income after square root will decrease by `r -summary(clean.nlsy.income2012.1.male.lm)$coef[4,1]` with a p value of `r summary(clean.nlsy.income2012.1.male.lm)$coef[4,4]`. There is no such significant relationship in female. **<br>
**3. Region. The income after square root of male in north east is `r summary(clean.nlsy.income2012.1.male.lm)$coef[5,1]` more than that of male in north central with a p value of `r summary(clean.nlsy.income2012.1.male.lm)$coef[5,4]`, while the income after square root of female in north east is `r summary(clean.nlsy.income2012.1.female.lm)$coef[5,1]` more than that of female in north central with a p value of `r summary(clean.nlsy.income2012.1.female.lm)$coef[5,4]`, so there is `r summary(clean.nlsy.income2012.1.male.lm)$coef[5,1] - summary(clean.nlsy.income2012.1.female.lm)$coef[5,1]` more difference. **<br>
**4. Education year. For male, every one more year of education, the income after square root will increase by `r summary(clean.nlsy.income2012.1.male.lm)$coef[8,1]` with a p value of `r summary(clean.nlsy.income2012.1.male.lm)$coef[8,4]`, while for female, every one more year of education, the income after square root will increase by `r summary(clean.nlsy.income2012.1.female.lm)$coef[8,1]` with a p value of `r summary(clean.nlsy.income2012.1.female.lm)$coef[8,4]`. So every one more year of education, the difference narrows by `r summary(clean.nlsy.income2012.1.female.lm)$coef[8,1] - summary(clean.nlsy.income2012.1.male.lm)$coef[8,1]`. **<br>
**5. Family size. For male, every one more person in family, the income after square root will increase by `r summary(clean.nlsy.income2012.1.male.lm)$coef[9,1]` with a p value of `r summary(clean.nlsy.income2012.1.male.lm)$coef[9,4]`, while for female, every one more person in family, the income after square root will decrease by `r -summary(clean.nlsy.income2012.1.female.lm)$coef[9,1]` with a p value of `r summary(clean.nlsy.income2012.1.female.lm)$coef[9,4]`. So every one more person in family, the difference increases by `r summary(clean.nlsy.income2012.1.male.lm)$coef[9,1] - summary(clean.nlsy.income2012.1.female.lm)$coef[9,1]`. **<br>
**6. Industry. Male in some of different industries has significant different income after square root, while there is no significant difference between female in different industries in income after square root. **<br><br>


<h1>Model 2</h1>
In the second model, we use the data of those whose income is more than 0, and create a logistic variable of whether the responder is of high income (whether the responder’s income is top coded), and do logistic regression.

```{r}
clean.nlsy.income2012.2 = select(cleaned.nlsy, 
                                 income.2012, 
                                 gender,
                                 race,  
                                 number.drink.per.day.2012, 
                                 residence.region.2012,
                                 highest.education.2012,
                                 family.size.2012,
                                 industry.2012
                                 ) %>%
  filter(income.2012>0)

for(i in 1:4663) {
  if(clean.nlsy.income2012.2[i, "income.2012"] == 343830) {
    clean.nlsy.income2012.2[i, "high.income"] = TRUE
  } else{
    clean.nlsy.income2012.2[i, "high.income"] = FALSE
  }
}

clean.nlsy.income2012.2 = select(clean.nlsy.income2012.2, -income.2012)

clean.nlsy.income2012.2.male = filter(clean.nlsy.income2012.2, gender == "Male") %>%
  select(-gender)
clean.nlsy.income2012.2.female = filter(clean.nlsy.income2012.2, gender == "Female")%>%
  select(-gender)
```


Regression summary of male: 
```{r}
clean.nlsy.income2012.2.male.glm = lm(high.income ~ ., 
                                      family = binomial(), 
                                      data = clean.nlsy.income2012.2.male)
summary(clean.nlsy.income2012.2.male.glm)$coef %>%
  kable() %>%
  kable_styling()
```

Regression summary of female: 
```{r}
clean.nlsy.income2012.2.female.glm = glm(high.income ~ ., 
                                         family = binomial(),
                                         data = clean.nlsy.income2012.2.female)
summary(clean.nlsy.income2012.2.female.glm)$coef %>%
  kable() %>%
  kable_styling()
```

<br>**Significant variables (with p value <= 0.05) of male are shown below: **

```{r}
summary(clean.nlsy.income2012.2.male.glm)$coef[which(
  summary(clean.nlsy.income2012.2.male.glm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```

<br>**Significant variables (with p value <= 0.05) of female are shown below: **

```{r}
summary(clean.nlsy.income2012.2.female.glm)$coef[which(
  summary(clean.nlsy.income2012.2.female.glm)$coef[, 4]<=0.05), ] 
```

This is education year. 

<br>**It is found that whether one will have top 2% income in 2012 vary depending on: **<br>
**1. Education year. For male, every one more year of education, the odds of having an income of top 2% increase by a factor of `r exp(summary(clean.nlsy.income2012.2.male.glm)$coef[8,1])` with a p value of `r summary(clean.nlsy.income2012.2.male.glm)$coef[8,4]`, while the odds of female is `r exp(summary(clean.nlsy.income2012.2.female.glm)$coef[8,1])` with a p value of `r summary(clean.nlsy.income2012.2.female.glm)$coef[8,4]`. The odds of female is `r exp(summary(clean.nlsy.income2012.2.female.glm)$coef[8,1]) - exp(summary(clean.nlsy.income2012.2.male.glm)$coef[8,1])` than male. **<br>
**2. Race. For male of other race compared to black, the odds of having top income increase by a factor of `r exp(summary(clean.nlsy.income2012.2.male.glm)$coef[3,1])` with a p value of `r summary(clean.nlsy.income2012.2.male.glm)$coef[3,4]` . There is no such significant relationship of female. **<br>
**3. Family size. For male, every one more person in family, the odds of having an income of top 2% increase by a factor of `r exp(summary(clean.nlsy.income2012.2.male.glm)$coef[9,1])` with a p value of `r summary(clean.nlsy.income2012.2.male.glm)$coef[9,4]`. There is no such significant relationship of female. **<br>
**4. Industry. Male in some of different industries has significant different possible of having top income, while there is no such significant difference between female. **<br><br>


<h1>Model 3</h1>
In the third model, we use every data and create a logistic variable of whether the responder has income (whether the income is more than 0), and do logistic regression.

```{r}
clean.nlsy.income2012.3 = select(cleaned.nlsy, 
                                 income.2012, 
                                 gender,
                                 race,  
                                 number.drink.per.day.2012, 
                                 residence.region.2012,
                                 highest.education.2012,
                                 family.size.2012,
                                 industry.2012
                                 ) 

for(i in 1:6021) {
  if(clean.nlsy.income2012.3[i, "income.2012"] > 0) {
    clean.nlsy.income2012.3[i, "have.income"] = TRUE
  } else{
    clean.nlsy.income2012.3[i, "have.income"] = FALSE
  }
}

clean.nlsy.income2012.3 = select(clean.nlsy.income2012.3, -income.2012)

clean.nlsy.income2012.3.male = filter(clean.nlsy.income2012.3, gender == "Male") %>%
  select(-gender)
clean.nlsy.income2012.3.female = filter(clean.nlsy.income2012.3, gender == "Female")%>%
  select(-gender)
```


Regression summary of male: 
```{r}
clean.nlsy.income2012.3.male.glm = lm(have.income ~ ., 
                                      family = binomial(), 
                                      data = clean.nlsy.income2012.3.male)
summary(clean.nlsy.income2012.3.male.glm)$coef %>%
  kable()
```

<br>
Regression summary of female: 
```{r}
clean.nlsy.income2012.3.female.glm = glm(have.income ~ ., 
                                         family = binomial(),
                                         data = clean.nlsy.income2012.3.female)
summary(clean.nlsy.income2012.3.female.glm)$coef %>%
  kable() %>%
  kable_styling()
```

<br>
**Significant variables (with p value <= 0.05) of male are shown below: **

```{r}
summary(clean.nlsy.income2012.3.male.glm)$coef[which(
  summary(clean.nlsy.income2012.3.male.glm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```

<br>
**Significant variables (with p value <= 0.05) of female are shown below: **

```{r}
summary(clean.nlsy.income2012.3.female.glm)$coef[which(
  summary(clean.nlsy.income2012.3.female.glm)$coef[, 4]<=0.05), ] %>%
  kable() %>%
  kable_styling()
```

<br>**It is found that whether one has no income in 2012 vary depending on: **<br>
**1. Race. For hispanic male compared to black male, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income2012.3.male.glm)$coef[2,1])` with a p value of `r summary(clean.nlsy.income2012.3.male.glm)$coef[2,4]`, but there is no such significant relationship for female. For other male compared to black male, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income2012.3.male.glm)$coef[3,1])` with a p value of `r summary(clean.nlsy.income2012.3.male.glm)$coef[3,4]`, while for other female compared to black female, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income2012.3.female.glm)$coef[3,1])` with a p value of `r summary(clean.nlsy.income2012.3.female.glm)$coef[3,4]`, so the difference narrows by `r exp(summary(clean.nlsy.income2012.3.female.glm)$coef[3,1]) - exp(summary(clean.nlsy.income2012.3.male.glm)$coef[3,1]) ` **<br>
**2. Times of drink per day. For male, every one more drink per day, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income2012.3.male.glm)$coef[4,1])` with a p value of `r summary(clean.nlsy.income2012.3.male.glm)$coef[4,4]`. For female, every one more drink per day, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income2012.3.female.glm)$coef[4,1])` with a p value of `r summary(clean.nlsy.income2012.3.female.glm)$coef[4,4]`. So, the difference narrows `r exp(summary(clean.nlsy.income2012.3.female.glm)$coef[4,1]) - exp(summary(clean.nlsy.income2012.3.male.glm)$coef[4,1])`**<br>
**3. Education year. For male, every one more year of education, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income2012.3.male.glm)$coef[8,1])` with a p value of `r summary(clean.nlsy.income2012.3.male.glm)$coef[8,4]`. For female, every one more year of education, the odds of having no income decrease by a factor of `r exp(summary(clean.nlsy.income2012.3.female.glm)$coef[8,1])` with a p value of `r summary(clean.nlsy.income2012.3.female.glm)$coef[8,4]`. So, the difference narrows `r exp(summary(clean.nlsy.income2012.3.female.glm)$coef[8,1]) - exp(summary(clean.nlsy.income2012.3.male.glm)$coef[8,1])`**<br>
**4. Industry.  Male in some of different industries has significant different possible of having no income, while there is no such significant difference between female. **<br><br>


<h1>Interaction model</h1>


The variabels needed are selected into `clean.nlsy.income2012.1`. For the income of 2012, the top 2% and zero income data is excluded.
```{r}
clean.nlsy.income2012.1 = select(cleaned.nlsy, 
                                 income.2012,
                                 gender,
                                 residence.region.2012,
                                 highest.education.2012,
                                 residence.urban.rural.2012,
                                 race,
                                 industry.2012,
                                 family.size.2012
                                 ) %>%
  filter(income.2012>0, 
         income.2012<max(cleaned.nlsy$income.2012))
```


Again, to test whether the income.2000 in clean.nlsy.income2000.1 is normally distributed, `qqmorm()` and `qqline()` are used here.
```{r}
par(mfrow=c(1,2))
qqnorm(clean.nlsy.income2012.1$income.2012^0.5)
qqline(clean.nlsy.income2012.1$income.2012^0.5)
qqnorm(clean.nlsy.income2012.1$income.2012)
qqline(clean.nlsy.income2012.1$income.2012)
```

Clearly, `income.2012^0.5` (left one) is more normally distributed than `income.2012` (right one). Therefore, `income.2012^0.5` is used in this interaction regression model.

Make each variable interact with `gender`. The summary and plots are as follows:
```{r}
income.lm.interact.2012 = lm(income.2012^0.5 ~
                               residence.region.2012*gender +
                               highest.education.2012*gender +  
                               family.size.2012*gender +  
                               residence.urban.rural.2012*gender + 
                               race*gender + 
                               industry.2012*gender, 
                             data = clean.nlsy.income2012.1)
summary(income.lm.interact.2012)
par(mfrow=c(2,2))
plot(income.lm.interact.2012)
```

Since there are several interaction terms that are not statistically significant, including  they are removed below to get a better interaction term model. And a new model is as follows:

```{r}
income.lm.interact.2012 = lm(income.2012^0.5 ~
                               highest.education.2012 +  
                               family.size.2012*gender + 
                               race*gender + 
                               industry.2012*gender, 
                             data = clean.nlsy.income2012.1)
summary(income.lm.interact.2012)
par(mfrow=c(2,2))
plot(income.lm.interact.2012)
```

There are three interaction terms indicate the significant association to the difference of income between male and female.

+ `family.size.2012:genderMale`: The coefficient of this term means that each time when the family size increases **one**, the difference of `income.2012^0.5` between male and female increases `r round(income.lm.interact.2012$coefficients["family.size.2012:genderMale"], 2)`.

+ `genderMale:raceOther`: the **Other race** significantly incluence the income difference between male and femal. In "Other" race, the **difference** of `income.2012^0.5` between male and femal increases `r round(income.lm.interact.2012$coefficients["genderMale:raceOther"], 2)`. Therefore, the income difference between male and femal in "Other" race will increase `r round((summary(income.lm.interact.2012)[[4]][1] + income.lm.interact.2012$coefficients["genderMale"] + income.lm.interact.2012$coefficients["raceOther"] + income.lm.interact.2012$coefficients["genderMale:raceOther"])^2 - (summary(income.lm.interact.2012)[[4]][1] + income.lm.interact.2012$coefficients["raceOther"])^2 - ((summary(income.lm.interact.2012)[[4]][1] + income.lm.interact.2012$coefficients["genderMale"])^2 - (summary(income.lm.interact.2012)[[4]][1])^2),2)`.

+ `genderMale:industry.2012Information`: in the **Information** industry, the difference of `income.2012^0.5` between male and female increases `r round(income.lm.interact.2012$coefficients["genderMale:industry.2012Information"], 2)`.

+ `genderMale:industry.2012Other services`: in **Other service** industry, the difference of `income.2012^0.5` between male and female increases `r round(income.lm.interact.2012$coefficients["genderMale:industry.2012Other services"], 2)`.

+ `genderMale:industry.2012Professional and Technical`: in **Professional and Technical** industry, the difference of `income.2012^0.5` between male and female increases `r round(income.lm.interact.2012$coefficients["genderMale:industry.2012Professional and Technical"], 2)`.



