---
title: "Final_Report_Part1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## 0. Introduction

> This project aims to analise the income difference between male and female and how did this income disparity varied depending on other factors. The dataset was meticulously gathered by questioning the same people on many different occasions across the span of thirty-three years. It was clear that there was income data from three separate years i.e. <b> 1990, 2000, and 2012.</b> And the report will study datas in all three years.
**Note: ** In order to make the report more readable, unless necessary, you may not see the all the code in html files. Please refer to Rmd files for the code.

------------

### Rename variables

In order to make the code more readable and descriptive, all the variables and factors used in the project are renamed before the analysis start.

```{r}
# Import starting data
nlsy <- read.csv("nlsy79_income.csv", header=TRUE)
```

```{r}
library("plyr")
library("dplyr")
library("ggplot2")
library("knitr")
library("kableExtra")
```


```{r}
# Change column names (mostly) to question name abbreviations
# Rename names of columns used in this report
colnames(nlsy) <- c("VERSION_R25_2012",
    "CASEID_1979",
    "FAM-2A_1979",
    "FAM-POB_1979",
    "FAM-3_1979",
    "FAM-3A_1979",
    "FAM-RES_1979",
    "FAM-6_1979",
    "R_REL-1_COL_1979",
    "SCHOOL-31_1979",
    "MIL-6_1979",
    "womens.roles.q1",
    "womens.roles.q2",
    "womens.roles.q3",
    "womens.roles.q4",
    "womens.roles.q6",
    "womens.roles.q7",
    "womens.roles.q8",
    "EXP-OCC_1979",
    "EXP-9_1979",
    "race",
    "gender",
    "marital.status.1979",
    "family.size.1979",
    "POVSTATUS_1979",
    "POLICE-1_1980",
    "POLICE-1C_1980",
    "POLICE-2_1980",
    "age.start.alcohol.1984",
    "times.drug.use.1984",
    "age.start.drug.use.1984",
    "income.1990",
    "POVSTATUS_1990",
    "HGCREV90_1990",
    "JOBSNUM_1990",
    "household.children.number.1990",
    "young.child.age.1990",
    "time.cocaine.use.1998",
    "age.start.drug.use.1998",
    "industry.2000",
    "CPSOCC80.01_2000",
    "OCCSP-55I_CODE_2000",
    "Q2-15B_2000",
    "Q10-2_2000",
    "income.2000",
    "family.size.2000",
    "TNFI_TRUNC_2000",
    "POVSTATUS_2000",
    "MARSTAT-COL_2000",
    "marital.status.2000",
    "MO1M1B_XRND",
    "Q2-10B~Y_2012",
    "industry.2012",
    "OCCALL-EMP.01_2012",
    "OCCSP-55I_CODE_2012",
    "Q2-15A_2012",
    "number.drink.per.day.2012",
    "income.2012",
    "Q13-5_SR000001_2012",
    "Q13-5_SR000002_2012",
    "Q13-18_TRUNC_2012",
    "Q13-18_SR000001_TRUNC_2012",
    "family.size.2012",
    "residence.region.2012",
    "highest.education.2012",
    "residence.urban.rural.2012",
    "JOBSNUM_2012")

# Map all negative values to missing (NA)
nlsy[nlsy < 0] <- NA

```

```{r}
# Transform relabel gender and race variables
nlsy = transform(nlsy, gender = as.factor(mapvalues(gender, 
                                                        c(1,2), 
                                                        c("Male", 
                                                          "Female"))))
nlsy = transform(nlsy, race = as.factor(mapvalues(race, 
                                                      c(1,2,3), 
                                                      c("Hispanic", 
                                                        "Black", 
                                                        "Other"))))
```

```{r}
# Transform the factors inside the residence.region.2012 variable to make it easier to understand
nlsy$residence.region.2012  = as.factor(mapvalues(nlsy$residence.region.2012 ,
                                                           c(1,2,3,4),
                                                           c("Northeast",
                                                             "North Central",
                                                             "South",
                                                             "West")))
```

```{r}
# Transform the factors inside the residence.urban.rural.2012 variable to make it easier to understand
nlsy$residence.urban.rural.2012  = as.factor(mapvalues(nlsy$residence.urban.rural.2012 ,
                                                           c(0,1,2),
                                                           c("Rural",
                                                             "Urban",
                                                             "Unknown"
                                                             )))
```

```{r}
# Transform the factors inside the marital.status.2000 variable to make it easier to understand
nlsy$marital.status.2000  = as.factor(mapvalues(nlsy$marital.status.2000 ,
                                                           c(0,1,2,3,6),
                                                           c("Never Married",
                                                             "Married",
                                                             "Separated",
                                                             "Divorced",
                                                             "Widowed")))
```

```{r}
# Modify and transform the industry.2000 variable to make it easier to analyse
for(i in 1:length(nlsy$industry.2000)) {
  if(is.na(nlsy[i, "industry.2000"])) {}
  else if(nlsy[i, "industry.2000"] >= 17
     && nlsy[i, "industry.2000"] <= 28) {
          nlsy[i, "industry.2000"] = 1
  }  else if(nlsy[i, "industry.2000"] >= 47
     && nlsy[i, "industry.2000"] <= 57) {
          nlsy[i, "industry.2000"] = 2
     } else if(nlsy[i, "industry.2000"] >= 67
     && nlsy[i, "industry.2000"] <= 77) {
          nlsy[i, "industry.2000"] = 3
     } else if(nlsy[i, "industry.2000"] >= 107
     && nlsy[i, "industry.2000"] <= 398) {
          nlsy[i, "industry.2000"] = 4
     } else if(nlsy[i, "industry.2000"] >= 407
     && nlsy[i, "industry.2000"] <= 479) {
          nlsy[i, "industry.2000"] = 5
     } else if(nlsy[i, "industry.2000"] >= 507
     && nlsy[i, "industry.2000"] <= 698) {
          nlsy[i, "industry.2000"] = 6
     } else if(nlsy[i, "industry.2000"] >= 707
     && nlsy[i, "industry.2000"] <= 718) {
          nlsy[i, "industry.2000"] = 7
     } else if(nlsy[i, "industry.2000"] >= 727
     && nlsy[i, "industry.2000"] <= 759) {
          nlsy[i, "industry.2000"] = 8
     } else if(nlsy[i, "industry.2000"] >= 769
     && nlsy[i, "industry.2000"] <= 798) {
          nlsy[i, "industry.2000"] = 9
     } else if(nlsy[i, "industry.2000"] >= 807
     && nlsy[i, "industry.2000"] <= 809) {
          nlsy[i, "industry.2000"] = 10
     } else if(nlsy[i, "industry.2000"] >= 828
     && nlsy[i, "industry.2000"] <= 897) {
          nlsy[i, "industry.2000"] = 11
     } else if(nlsy[i, "industry.2000"] >= 907
     && nlsy[i, "industry.2000"] <= 937) {
          nlsy[i, "industry.2000"] = 12
     }else{
       nlsy[i, "industry.2000"] = 13
     }
}

nlsy$industry.2000 = as.factor(
  mapvalues(
    nlsy$industry.2000,
    c(1,2,3,4,5,6,7,8,9,10,11,12,13),
    c("AGRICULTURE,FORESTRY AND FISHERIES",
      "MINING",
      "CONSTRUCTION",
      "MANUFACTURING",
      "TRANSPORTATION,COMMUNICATION,PUBLIC UTILITIES",
      "WHOLESALE AND RETAIL TRADE",
      "FINANCE,INSURANCE AND REAL ESTATE",
      "BUSINESS AND REPAIR SERVICES",
      "PERSONAL SERVICES",
      "ENTERTAINMENT AND RECREATION SERVICES",
      "PROFESSIONAL AND RELATED SERVICES",
      "PUBLIC ADMINISTRATION",
      "OTHERS")))

```

```{r}
# modify the industry.2012 variable to make it easier to analyse
for(i in 1:length(nlsy$industry.2012)) {
  if(is.na(nlsy[i, "industry.2012"])) {}
  else if(nlsy[i, "industry.2012"] >= 170
     && nlsy[i, "industry.2012"] <= 290) {
          nlsy[i, "industry.2012"] = 1
  }  else if(nlsy[i, "industry.2012"] >= 370
     && nlsy[i, "industry.2012"] <= 490) {
          nlsy[i, "industry.2012"] = 2
     } else if(nlsy[i, "industry.2012"] >= 570
     && nlsy[i, "industry.2012"] <= 690) {
          nlsy[i, "industry.2012"] = 3
     } else if(nlsy[i, "industry.2012"] == 770) {
          nlsy[i, "industry.2012"] = 4
     } else if(nlsy[i, "industry.2012"] >= 1070
     && nlsy[i, "industry.2012"] <= 3990) {
          nlsy[i, "industry.2012"] = 5
     } else if(nlsy[i, "industry.2012"] >= 4070
     && nlsy[i, "industry.2012"] <= 4590) {
          nlsy[i, "industry.2012"] = 6
     } else if(nlsy[i, "industry.2012"] >= 4670
     && nlsy[i, "industry.2012"] <= 5790) {
          nlsy[i, "industry.2012"] = 7
     } else if(nlsy[i, "industry.2012"] >= 6070
     && nlsy[i, "industry.2012"] <= 6390) {
          nlsy[i, "industry.2012"] = 8
     } else if(nlsy[i, "industry.2012"] >= 6470
     && nlsy[i, "industry.2012"] <= 6780) {
          nlsy[i, "industry.2012"] = 9
     } else if(nlsy[i, "industry.2012"] >= 6870
     && nlsy[i, "industry.2012"] <= 6990) {
          nlsy[i, "industry.2012"] = 10
     } else if(nlsy[i, "industry.2012"] >= 7070
     && nlsy[i, "industry.2012"] <= 7190) {
          nlsy[i, "industry.2012"] = 11
     } else if(nlsy[i, "industry.2012"] >= 7270
     && nlsy[i, "industry.2012"] <= 7490) {
          nlsy[i, "industry.2012"] = 12
     } else if(nlsy[i, "industry.2012"] >= 7570
     && nlsy[i, "industry.2012"] <= 7790) {
          nlsy[i, "industry.2012"] = 13
     } else if(nlsy[i, "industry.2012"] >= 7860
     && nlsy[i, "industry.2012"] <= 7890) {
          nlsy[i, "industry.2012"] = 14
     } else if(nlsy[i, "industry.2012"] >= 7970
     && nlsy[i, "industry.2012"] <= 8470) {
          nlsy[i, "industry.2012"] = 15
     } else if(nlsy[i, "industry.2012"] >= 8560
     && nlsy[i, "industry.2012"] <= 8590) {
          nlsy[i, "industry.2012"] = 16
     } else if(nlsy[i, "industry.2012"] >= 8660
     && nlsy[i, "industry.2012"] <= 8690) {
          nlsy[i, "industry.2012"] = 17
     } else if(nlsy[i, "industry.2012"] >= 8770
     && nlsy[i, "industry.2012"] <= 9290) {
          nlsy[i, "industry.2012"] = 18
     } else if(nlsy[i, "industry.2012"] >= 9370
     && nlsy[i, "industry.2012"] <= 9870) {
          nlsy[i, "industry.2012"] = 19
     } else if(nlsy[i, "industry.2012"] == 9890) {
          nlsy[i, "industry.2012"] = 20
     } else if(nlsy[i, "industry.2012"] >= 9950) {
          nlsy[i, "industry.2012"] = 21
     } else if(nlsy[i, "industry.2012"] >= 9990) {
          nlsy[i, "industry.2012"] = 22
     }  else {
          nlsy[i, "industry.2012"] = 25
     }
}

nlsy$industry.2012 = as.factor(
  mapvalues(
    nlsy$industry.2012,
    c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,25),
    c("Agriculture, Forestry, Fishing, and Hunting",
      "Mining",
      "Utilities",
      "Construction",
      "Manufacturing",
      "Wholesale Trade",
      "Retail Trade",
      "Transportation and Warehousing",
      "Information",
      "Finance and Insurance",
      "Real Estate",
      "Professional and Technical",
      "Administration and Management",
      "Education",
      "Health Care",
      "Arts and Entertainment",
      "Accomodation and Food",
      "Other services",
      "Public Administration",
      "Armed Forces",
      "Not in Labor Force",
      "Uncodeable",
      "Uncategorized")))
```

```{r}
# Transform womens.roles.q7 variable
nlsy$womens.roles.q7 = as.factor(mapvalues(nlsy$womens.roles.q7, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q1 variable
nlsy$womens.roles.q1 = as.factor(mapvalues(nlsy$womens.roles.q1, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q2 variable
nlsy$womens.roles.q2 = as.factor(mapvalues(nlsy$womens.roles.q2, 
                                                    c(1,2,3,4), 
                                                    c("Strongly disagree",
                                                      "Disagree", 
                                                      "Agree", 
                                                      "Strongly agree")))

```

```{r}
# Transform womens.roles.q4 variable
nlsy$womens.roles.q4 = as.factor(mapvalues(nlsy$womens.roles.q4, 
                                           c(1,2,3,4), 
                                           c("Strongly disagree", 
                                             "Disagree", 
                                             "Agree", 
                                             "Strongly agree")))
```


Notice that it is only a preliminary data processing, more detailed data handeling process will be displayed in the **Methodology** part.

## 1. Data Summary

> This part selects a small part of variables considered to be significant variables associated with the income disparity by project team at first. However, some of them may not be further analysed in the following part due to insignificance or colinearity.

---------------

### Family Attitude towards Women's Role 
The respondents were asked the question "EMPLOYMENT OF WIVES LEADS TO JUVENILE DELINQUENCY?" about their family attitudes during the year 1979. This attitude was indicative of an ultra-conservative mindset and hence it was chosen as an indicator for gender income disparity. The below table shows the variation of income based on the answer to the question.
```{r}
nlsy$womens.roles.q4 = factor(nlsy$womens.roles.q4, 
                                               order = TRUE, 
                                               levels = c("Strongly disagree", 
                                                          "Disagree", 
                                                          "Agree", 
                                                          "Strongly agree"))
```

```{r}
ddply(nlsy, 
      c("womens.roles.q4", "gender"), 
      summarise, 
      mean.income = mean(income.1990, na.rm=TRUE), 
      sd.income = sd(income.1990, na.rm=TRUE), 
      median.income = median(income.1990, na.rm=TRUE)
      ) %>%
kable(col.names = c("Response",
                    "gender",
                    "Mean income",
                    "Standard deviation of income",
                    "Median income"),
      align = rep('c',5)) %>%
  kable_styling()
```

```{r}
Womens.roles.4.summary = aggregate(income.1990 ~ womens.roles.q4 + 
                                     gender, 
                                   data = nlsy, FUN = mean)
```

Here's a graph showing the male and female income gropued together based on the responses: 
```{r}
ggplot(Womens.roles.4.summary, aes(x=womens.roles.q4, 
                                   y=income.1990, fill=gender)) +
  geom_bar(stat="identity", position="dodge") +
  xlab("EMPLOYMENT OF WIVES LEADS TO JUVENILE DELINQUENCY?") +
  ylab("Income in the year 1990")
```
It is clear that the individuals who have this narrow-minded mentality have lower incomes irrespective of gender. But the disparity within the gender is unclear from this graph.

-------------------
### Drug Use - Cocaine

In 1998, the respondents were questioned about the number of times they used cocain in their lifetime.
```{r}
ddply(nlsy, 
      "gender", summarise, 
      num.drug.used = length(time.cocaine.use.1998[which((
        !is.na(time.cocaine.use.1998))&(time.cocaine.use.1998>0))]), 
      mean.drug.time = mean(time.cocaine.use.1998, na.rm=TRUE), 
      sd.drug.time = sd(time.cocaine.use.1998, na.rm=TRUE), 
      median.drug.time = median(time.cocaine.use.1998, na.rm=TRUE))%>%
kable(col.names = c("gender",
                    "Number of times used cocaine (N)",
                    "Mean of N",
                    "Standard deviation of N",
                    "Median of N"),
      align = rep('c',5)) %>%
  kable_styling()
```

It is clear that the most common response is 0 times i.e. many people have never used cocaine in their life. But amogst those who have, men are more prone to use.

```{r}
ggplot(nlsy, 
       aes(x=time.cocaine.use.1998, y=income.2000, color=gender)) +
  geom_point() +
  facet_wrap(~ gender) +
  geom_smooth(method="lm") +
  xlab("How many times use drug in 1998") +
  ylab("Income in 2000")
```

The factor might seem important in terms of affecting the income of an individual but it affects male income slightly and doesnt affect female income much. This could be because men are more prone to using cocaine in the first place.

---------------

### Number of children in the household

The number of biological or step or adopted children in the house seems relevant to the income of an individual. The data about the number of children in the household in 1990 was used to relate it with the income in 1990.

```{r}
ddply(nlsy, 
      "gender", summarise, 
      num.child = length(household.children.number.1990), 
      mean.num.child = mean(household.children.number.1990, na.rm=TRUE), 
      sd.num.child = sd(household.children.number.1990, na.rm=TRUE), 
      median.num.child = median(household.children.number.1990, na.rm=TRUE)) %>%
kable(col.names = c("gender",
                    "Total of number of children in the household (N)",
                    "Mean of N",
                    "Standard deviation of N",
                    "Median of N"),
      align = rep('c',5)) %>%
  kable_styling()
```

From the table, it is clear that the average number of children in a female's household is more than a male's.

```{r}
ggplot(nlsy, 
       aes(x=household.children.number.1990, y=income.2000, color=gender)) +
  geom_point() +
  facet_wrap(~ gender) +
  geom_smooth(method="lm") +
  xlab("Number of bio/step/adpt children in household in 1990") +
  ylab("Income in 2000")
```

For women, the more children they have in the household, the lower income they earn, while there is no such a trend for men. In fact there is a slight increase in male income with the increase in the number of children in the household. Women are expected to take care of the household and increase in the number of kids affects their ability to earn tremendously. 


--------------------

### Age of youngest child

The age of the youngest child is a similar variable which seems to have some relevance in this regard.

```{r}
ddply(nlsy, 
      "gender", summarise, 
      mean.age.youngch = mean(young.child.age.1990, na.rm=TRUE), 
      sd.age.youngch = sd(young.child.age.1990, na.rm=TRUE), 
      median.age.youngch = median(young.child.age.1990, na.rm=TRUE)) %>%
kable(col.names = c("gender",
                    "Mean age of youngest child",
                    "Standard deviation of the age of youngest child",
                    "Median of the age of youngest child"),
      align = rep('c',5)) %>%
  kable_styling()
```

The youngest child's age in female respondent's household is on average greater than males respondent's household.
The above data can be used to correlated with the income as follows:

```{r}
ggplot(nlsy, 
       aes(x=young.child.age.1990, y=income.1990, color=gender)) +
  geom_smooth() + 
  ylab("Income in 1990") +
  xlab("Age of youngest bio/step/adpt child in household in 1990")
```

Women with younger kids tend to earn less. Especially if the age of the youngest child is between 0-5 female income is adversely affected. Male income on the other hand dips slightly after the teen years of the child.


--------------------

### Region

For 2012, Region from which the respondent hails from seems important Income Mapping. Here is a comparison of incomes based on region and gender:

```{r}
region.2012.summary = aggregate(income.2012 ~ residence.region.2012 + gender, 
                                   data = nlsy, FUN = mean)

ggplot(region.2012.summary, aes(x=residence.region.2012, y=income.2012, fill=gender)) +
  geom_bar(stat="identity", position="dodge") +
  xlab("REGION")
```

Clearly there is a gap between male and female incomes in each of the four regions. But this can be due to a variety of factors which we explore in our methodlogy and findings. 

--------------------

### Urban/Rural

For the 2012 income data, a similar trend can be seen in the urban and rural population.

```{r}
ggplot(nlsy, aes(x=residence.urban.rural.2012, y=income.2012 ,fill=gender)) +
  geom_boxplot() +
  xlab("Living in Urban Area or Rural?")
```

The male income is greater the femal income in almost all regions. A quartile of the population has zero income in all regions, except the male population in rural or urban areas. 


--------------------

### Race, Gender and Income

A distribution of income in 2012 by a combination of gender and race is given below. 

```{r}
ddply(nlsy, 
      c("race", "gender"), 
      summarise, 
      mean.income = mean(income.2012, na.rm=TRUE), 
      sd.income = sd(income.2012, na.rm=TRUE), 
      median.income = median(income.2012, na.rm=TRUE)
      )%>%
kable(col.names = c("Race",
                    "gender",
                    "Mean income",
                    "Standard deviation of income",
                    "Median income"),
      align = rep('c',5)) %>%
  kable_styling()
```

The income varies greately across the three races and two genders with the 'Other' race and 'male' gender having greater income than other races and gender. 

```{r}
ggplot(nlsy, aes(x=race, fill=gender)) +
  geom_bar(stat="count", position="dodge") +
  xlab("RACE")
```
This difference in income in comparison comes with a large number of people in the 'Other' race which makes this difference even more significant. 

```{r}
race.summary = aggregate(income.1990 ~ race + gender, 
                         data = nlsy, FUN = mean)
```

```{r}
ggplot(race.summary, aes(x=race, y=income.1990, fill=gender)) +
  geom_bar(stat="identity", position="dodge") +
  xlab("RACE")
```
From a combined income comparison plot the income disparity is clear and obvious. But we need to explore the other data variables to get a fair assessment of gender bias in terms of income for male and female respondents of the survey. 










